A Typed Programmatic Interface to Contracts on the Blockchain

### 2021-06-14 papers/execution-model/APLAS 2021/paper.tex
snapshot-section3-types

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{mathpartir}
\usepackage{fixme}
\usepackage{xcolor}
\usepackage{listings}
\lstdefinelanguage{michelson}{
%  basicstyle=\ttfamily,
  morekeywords={parameter,storage,or,unit,mutez,pair,bool,address}, sensitive=false,
  morecomment=[l]{\#},
  morecomment=[s]{/*}{*/},
  morestring=[b]",
}
\lstset{language=Caml}
\input{macros}

\begin{document}
%
\title{A Typed Programmatic Interface to Contracts on the Blockchain}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Thi Thu Ha Doan\orcidID{0000-0001-7524-4497}\and
Peter Thiemann\orcidID{0000-0002-9000-1239}}

%
\authorrunning{~}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of Freiburg, Germany \\
\email{\{doanha,thiemann\}@informatik.uni-freiburg.de}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
  Smart contract applications on the blockchain can only reach their full potential if
  they can be integrated with traditional software systems. Invoking
  contracts and observing the state of the blockchain requires a
  programmatic interface. We propose a typed API for this
  purpose and establish some properties of the combined
  system. Specifically, we provide an execution model that
  enables us to prove type-safe interaction between programs and the 
  blockchain. We establish further properties of the model that
  give rise to requirements on the API. A prototype of the interface
  is implemented in OCaml for the Tezos blockchain.
\end{abstract}

@@

In Tezos, there are two types of accounts: implicit accounts, which are associated with a pair of private/public keys, and smart contracts, which are programmable accounts created by an origination operation and have no corresponding secret key. The address of a smart contract is a unique public hash that depends on the creation operation. An implicit account is maintained on the blockchain with its public key, balance, and counter, whose job is to prevent replay attacks. A smart contract account is stored with its script, storage, and balance. A contract script maps a pair of a parameter and a storage, which have fixed and monomorphic types to a pair of a list of internal operations and an updated storage. An implicit account performs three kinds of transactions: (1) transfer tokens to another implicit account, (2) invoke a smart contract, or (3) originate a new smart contract. A contract origination specifies the script of the contract and the initial contents of the contract storage, while a contract invocation must provide input data. Each transaction contains a fee to be paid either by payment to a baker or by destruction (burning). A transaction is injected into the blockchain network via a node, which then validates the transaction before submitting it to the network. Generally, a transaction can be rejected by the node for a number of reasons. Once the transaction is valid, it is injected into the mempool, which contains all pending transactions before they can be included in a block. On the other hand, a pending transaction can be lost in the mempool. A transaction is timed out when 60 blocks have passed and it can no longer be included in a block. When a transaction is included in a blockchain, the related accounts are updated according to the transaction result.

The execution model is an abstraction of the Tezos blockchain \cite{tezos-whitepaper}
augmented with functional (OCaml) programs that interact with the
blockchain. As the blockchain is realized by a peer-to-peer network of
independent nodes, interaction happens through
\emph{local nodes} that receive requests to originate and invoke
contracts from programs that run on a particular node. We model the blockchain
itself as a separate, abstract global entity that represents the
current consensual state of the system. Our model does not express
low-level details, but relies on nondeterminism to describe
the possible behaviors of the system. In particular, we do
\textbf{not} formalize the execution of the smart contracts
themselves, we rather consider them as black boxes that execute in a
safe way. Tezos' smart contract language Michelson and its
properties have been formalized elsewhere \cite{DBLP:conf/fm/BernardoCHPT19}.

@@

The local node offers operations to transfer tokens from one account
to another, to invoke a contract, and to originate a contract on the blockchain.
\begin{align*}
  \OP &::= \TRANSFER[\PARAMETER]\NTEZ\PUK\ADDR\MTEZ
  \\&\mid \ORIGINATE\NTEZ\PUK\CODE\INIT\MTEZ
\end{align*}
In the transfer, which also serves as invocation, \NTEZ\ is the amount of tokens transferred, \PUK\ is the public key of the
sender, \ADDR\ is either a public 
key for an implicit account (in case of a simple transfer)  or a
public hash for a smart contract (for an invocation), \PARAMETER\ is
the argument passed to the smart contract, 
which is empty for a simple transfer, and \MTEZ\ is
the amount of tokens for the transaction fee. In originate,  \CODE\ is
the script of a smart contract and \INIT\ is the initial value of the
contract's storage. Each operation returns an \emph{operation hash}
\OPH, on which we can query the status of the operation.

The local node offers several ways to query the current state of
the blockchain.
\emph{Query operators} are defined by the following grammar:
\begin{align*}
  \QOP & ::=
  % \GETCOUNTER\PUK  
  %        \mid
         \GETBALANCE{}%\ADDR
  \mid \GETSTATUS{}%\OPH
  \mid \GETSTORAGE{}%\PUH
  % \mid \GETCODE\PUH 
  % \mid \GETTYPE\PUH
  \mid \GETCONTRACT{}%\OPH
\end{align*}
We obtain the balance associated with an implicit account or a
contract by its public key or public hash, respectively; the status of
a submitted operation by its operation hash; the stored value of a
contract by its public hash; and the public hash of a contract by the
operation hash of its originating transaction.

@@

\begin{figure}[tp]
\begin{align*}
  \CONSTANT & ::= \INT \mid \FIX
              \mid \OPH \mid \PUH \mid\PUK \mid \CODE \mid
              \NTEZ \mid \SUNIT \mid \FALSE \mid \TRUE \\
  \STATUS & ::= \STATUSPENDING \mid \STATUSINCLUDING (\INT) \mid
            \STATUSTIMEOUT \\
  \ERROR &::= \ERRPRG \mid \ERRBAL \mid \ERRCOUNT \mid \ERRFEE \mid
           \ERRPUK \mid \ERRPUH \mid \ERRARG \mid \ERRINIT\\
	\EXPR & :: =  \CONSTANT \mid \STATUS \mid \ERROR \mid \VARIABLE \mid \lambda \VARIABLE. \EXPR\ 
	\mid \EXPR\EXPR  %\mid \FST\ \EXPR \mid \SND\ \EXPR 	%\mid \NI \EI
  \mid \EXPR\ \PLUS\ \EXPR\ \mid
  % \EXPR\ \MINUS\ \EXPR\ \mid
  \EXPR\ \EQUAL\ \EXPR\
  % \mid  \EXPR\ \LESS\ \EXPR
  \mid \EXPR\ \AND\ \EXPR\ \mid \EXPR\ \OR\ \EXPR\ \mid \NOT\ \EXPR
  \\& \mid (\EXPR,\EXPR)\mid \NIL\ \mid \CONS\ \EXPR\ \EXPR \mid \LEFT\ \EXPR \mid\RIGHT\
  \EXPR \mid \SOME\ \EXPR \mid \NONE
  \mid \MATCH\ \EXPR\ \WITH\ \PATTERN\to\EXPR \dots 
  \\& \mid \RAISE\ \EXPR \mid \TRY\ \EXPR\ \EXCEPT\ \EXPR \mid \CAST\EXPR\TYPE\TYPEU
  \\& \mid \QOP\  \EXPR \mid \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR
  \\& \mid \ORIGINATE\EXPR\EXPR\EXPR\EXPR\EXPR
  % \\&\mid \CONTRACT\ \TYPE\ \EXPR
  \\& \mid \TOSTRING\ \EXPR \mid \FROMSTRING\ \TYPE\ \EXPR
  \\
  \PATTERN &::= \VARIABLE \mid (\PATTERN, \PATTERN) \mid \NIL \mid \CONS\ \PATTERN\ \PATTERN \mid \LEFT\ \PATTERN \mid\RIGHT\
             \PATTERN \mid \SOME\ \PATTERN \mid \NONE \\
            &\mid \FALSE \mid \TRUE \mid \STATUS \mid \ERROR
  % \\
  % \QOP & ::= \GETBALANCE{} \mid \GETSTATUS{} \mid \GETSTORAGE{} \mid \GETCODE{} \mid \GETCOUNTER{}
  %        \mid \GETTYPE{}
  \\[2ex]
  \TYPE, \TYPEU & ::=
                  \TPUH \mid
                  \TPUK \mid
                  \TADDR \mid 
                  \TCONTRACT\ \TYPE \mid
                  \TCODE\ \TYPE\ \TYPEU \mid \TTEZ \mid
                  \TOPH \mid
                  \TSTATUS \mid \TEXCEPTION \\
  & \mid \TINT \mid \TUNIT \mid \TBOOL \mid \TSTRING \mid \TYPE\to\TYPEU \mid \TPAIR\ \TYPE\ \TYPEU \mid \TLIST\ \TYPE
    \mid \TSUM\ \TYPE\ \TYPEU \mid \TOPTION\ \TYPE 
\end{align*}
  \caption{Syntax of expressions, \EXPR, and types, \TYPE}
  \label{fig:syntax-expressions}
\end{figure}
All queries and operations are started by programs, i.e., closed
expressions of type unit that run on the
local node. Each program can send transactions on behalf of any
account on the local node.  Figure~\ref{fig:syntax-expressions} defines the syntax of
expressions, which describe the usual constructs of call-by-value PCF
with sum, product, list, and option types, exceptions and fixpoint. Moreover, there are
domain-specific primitive types and constants \CONSTANT\ to support blockchain
operations, as well as several exceptional values collected in \ERROR.
Pattern matching is the only means to decompose values, cf.\ \PATTERN.
There is syntax to initiate transfers and to originate contracts as
well as for the queries. Finally, there is a type cast
$\CAST\EXPR\TYPE\TYPEU$, which we describe after discussing types.

Types (also in Figure~\ref{fig:syntax-expressions}) comprise a
collection of standard base types as well as 
functions, pairs, lists, sums, and option types. In addition, there
are domain specific types of public hashes $\TPUH$ and public keys $\TPUK$ subsumed by
a type of addresses $\TADDR$. $\TCONTRACT\ \TYPE$ is the type of
a contract with parameter type $\TYPE$. $\CODE\ \TYPE\ \TYPEU$
indicates a Michelson program with parameter type $\TYPE$ and storage
type $\TYPEU$. Tezos tokens have type $\TTEZ$. The type $\TOPH$ signifies operation hashes
returned by blockchain operations. Querying the status of an operation
returns a value of type $\TSTATUS$. Exceptions have type
$\TEXCEPTION$.

@@

Type casts are only applicable to certain pairs of
types governed by a relation $\SubType$, which could also serve as a
subtyping relation. It is given by the axioms
$\inferrule{}{\TPUH \SubType \TADDR}$, $\inferrule{}{\TPUK \SubType
  \TADDR}$, and $\inferrule{}{\TCONTRACT\ \TYPE \SubType \TPUH}$.
A cast from $\TYPE$ to $\TYPEU$ is only allowed
if $\TYPE \SubType \TYPEU$ (upcast) or $\TYPEU \SubType \TYPE$
(downcast). Upcasts always succeed, but downcasts may fail at run time.
In particular,  public hashes and public keys can both stand for addresses. Moreover, a smart
contract with parameter type $\TYPE$ is represented by its public hash
at run time. The corresponding downcast must check whether the public
hash is valid and has the expected parameter type.

@@

    \inferrule{}{\JTypeExpr\TEnv\OPH\TOPH}

@@

    \inferrule{
      \JTypeExpr\TEnv\EXPR\TYPE \\
      \TYPE \SubType \TYPEU \vee \TYPEU \SubType \TYPE
    }{
      \JTypeExpr\TEnv{\CAST\EXPR\TYPE\TYPEU}\TYPEU
    }
  \end{mathpar}
  \caption{Typing rules for expressions (excerpt)}
  \label{fig:typing-expressions}
\end{figure}
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{
      \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
      \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
      \JTypeExpr\TEnv{\EXPR_3}\TPUK \\
      % \JTypeExpr\TEnv{\EXPR_3}{\TCONTRACT\ \TUNIT} \\
      \JTypeExpr\TEnv{\EXPR_4}\TUNIT \\
      \JTypeExpr\TEnv{\EXPR_5}\TTEZ }{
      \JTypeExpr\TEnv{\TRANSFER[\EXPR_4]{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_5}}\TOPH
    }

    \inferrule{
      \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
      \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
      \JTypeExpr\TEnv{\EXPR_3}\TCONTRACT\ \TYPE \\
      \JTypeExpr\TEnv{\EXPR_4}{\TYPE} \\
      \JTypeExpr\TEnv{\EXPR_5}\TTEZ }{
      \JTypeExpr\TEnv{\TRANSFER[\EXPR_4]{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_5}}\TOPH
    }

    \inferrule{
      \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
      \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
      \JTypeExpr\TEnv{\EXPR_3}{\TCODE\ \TYPE_p\ \TYPE_s} \\
      \JTypeExpr\TEnv{\EXPR_4}\TYPE_s \\ % really \TSTRING \\
      \JTypeExpr\TEnv{\EXPR_5}\TTEZ }{
      \JTypeExpr\TEnv{\ORIGINATE{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_4}{\EXPR_5}}\TOPH
    }
  \end{mathpar}
\begin{mathpar}
  % \inferrule{
  %   \JTypeExpr\TEnv\EXPR\TPUK
  % }{
  %   \JTypeExpr\TEnv {\GETCOUNTER\EXPR}\TINT
  % }
  %   
  \inferrule{
    \JTypeExpr\TEnv\EXPR\TADDR
  }{
    \JTypeExpr\TEnv {\GETBALANCE\EXPR}\TTEZ
  }

  \inferrule{
    \JTypeExpr\TEnv\EXPR\TOPH
  }{
    \JTypeExpr\TEnv {\GETSTATUS\EXPR}\TSTATUS
  }

  \inferrule{
    \JTypeExpr\TEnv\EXPR{\TPUH}
  }{
    \JTypeExpr\TEnv {\GETSTORAGE\EXPR}\TSTRING
  }

  % \inferrule{
  %   \JTypeExpr\TEnv\EXPR{\CONTRACT\ \TYPE}
  % }{
  %   \JTypeExpr\TEnv {\GETCODE\EXPR}?
  % }
  %   
  % \inferrule{
  %   \JTypeExpr\TEnv\EXPR{\CONTRACT\ \TYPE}
  % }{
  %   \JTypeExpr\TEnv {\GETTYPE\EXPR}?
  % }
  %   
  \inferrule{
    \JTypeExpr\TEnv\EXPR{\TOPH}
  }{
    \JTypeExpr\TEnv {\GETCONTRACT\EXPR}\TPUH
  }
\end{mathpar}
  \caption{Typing rules for blockchain operations and queries}
  \label{fig:typing-blockchain-operations}
\end{figure}
% Generic rule for transfers
% \begin{mathpar}
%  \inferrule{
%    \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
%    \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
%    \JTypeExpr\TEnv{\EXPR_3}\TADDR \\
%    \JTypeExpr\TEnv{\EXPR_4}\TSTRING \\
%    \JTypeExpr\TEnv{\EXPR_5}\TTEZ
%  }{
%    \JTypeExpr\TEnv{\TRANSFER[\EXPR_4]{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_5}}\TOPH
%  }
% \end{mathpar}

Conversion to and from serialized storage; deserialization may fail
with an exception
\begin{mathpar}
  \inferrule{
    \JTypeExpr\TEnv\EXPR\TYPE \\
    \SERIALIZABLE (\TYPE)
  }{
    \JTypeExpr\TEnv{\TOSTRING\ \EXPR}\TSTRING
  }

  \inferrule{
    \JTypeExpr\TEnv{\EXPR}\TSTRING
  }{
    \JTypeExpr\TEnv{\FROMSTRING\ \TYPE\ \EXPR}\TYPE
  }
\end{mathpar}

>>>>

% APLAS 2021: Regular research papers should not exceed 18 pages in
% the Springer LNCS format(LaTeX template), including bibliography and
% figures.
% Lightweight double-blind: Author names and institutions must be
% omitted and References to the authors’ own related work should be in
% the third person 
%
\documentclass[runningheads]{llncs}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{mathpartir}
\usepackage{fixme}
\usepackage{xcolor}
\usepackage{listings}
\lstdefinelanguage{michelson}{
%  basicstyle=\ttfamily,
  morekeywords={parameter,storage,or,unit,mutez,pair,bool,address}, sensitive=false,
  morecomment=[l]{\#},
  morecomment=[s]{/*}{*/},
  morestring=[b]",
}
\lstset{language=Caml}
\input{macros}

\begin{document}
%
\title{A Typed Programmatic Interface to Contracts on the Blockchain}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
% ANONYMOUS
% \author{Thi Thu Ha Doan\orcidID{0000-0001-7524-4497}\and
% Peter Thiemann\orcidID{0000-0002-9000-1239}}

%
\authorrunning{~}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
% ANONYMOUS
% \institute{University of Freiburg, Germany \\
% \email{\{doanha,thiemann\}@informatik.uni-freiburg.de}
% }
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
  Smart contract applications on the blockchain can only reach their full potential if
  they can be integrated with traditional software systems. Invoking
  contracts and observing the state of the blockchain requires a
  programmatic interface. We propose a typed API for this
  purpose and establish some properties of the combined
  system. Specifically, we provide an execution model that
  enables us to prove type-safe interaction between programs and the 
  blockchain. We establish further properties of the model that
  give rise to requirements on the API. A prototype of the interface
  is implemented in OCaml for the Tezos blockchain.
\end{abstract}

@@

Tezos supports two types of accounts: implicit accounts, which
are associated with a pair of private/public keys, and smart
contracts, which are programmable accounts created by an origination
operation. The address of a smart
contract is a unique public hash that depends on the creation
operation. No key pair is associated with a smart contract. An
implicit account is maintained on the blockchain with 
its public key and balance.
% , and counter, whose job is to preventreplay attacks.
A smart contract account is stored with its script, storage,
and balance. A contract script maps a pair of a parameter and a
storage, which have fixed and monomorphic types, to a pair of a list of
internal operations and an updated storage. An account can
perform three kinds of transactions: (1) transfer tokens to an
implicit account, (2) invoke a smart contract, or (3) originate a new
smart contract. A contract origination specifies the script of the
contract and the initial contents of the contract storage, while a
contract invocation must provide input data. Each transaction contains
a fee to be paid either by payment to a baker or by destruction
(burning). A transaction is injected into the blockchain network via a
node, which then validates the transaction before submitting it to the
network. A transaction may be rejected by the node for a
number of reasons. After validation, the transaction is injected into
a \emph{mempool}, which contains all pending transactions before they can
be included in a block. A pending transaction may simply disappear
from the mempool, for example, a transaction times out when 60 blocks
have passed and it can no longer be included in a block. When a transaction
is included in the blockchain, the affected accounts are updated
according to the transaction result.

The execution model consists of functional (OCaml) programs that
interact with an abstraction of the Tezos blockchain
\cite{tezos-whitepaper}. As the blockchain is realized by a
peer-to-peer network of 
independent nodes, interaction happens through
\emph{local nodes} that receive requests to originate and invoke
contracts from programs that run on a particular node. We model the blockchain
itself as a separate, abstract global entity that represents the
current consensual state of the system. Our model does not express
low-level details, but relies on nondeterminism to describe
the possible behaviors of the system. In particular, we do
\textbf{not} formalize the execution of the smart contracts
themselves, we rather consider them as black boxes and probe their
observable behavior. Tezos' smart contract language Michelson and its
properties have been formalized elsewhere \cite{DBLP:conf/fm/BernardoCHPT19}.

@@

The local node offers operations to transfer tokens from one account
to another, to invoke a contract, and to originate a contract on the blockchain.
\begin{align*}
  \OP &::= \TRANSFER[\PARAMETER]\NTEZ\PUK\ADDR\MTEZ
  \\&\mid \ORIGINATE\NTEZ\PUK\CODE\INIT\MTEZ
\end{align*}
In the transfer, which also serves as contract invocation, \NTEZ\ is the amount of tokens transferred, \PUK\ is the public key of the
sender, \ADDR\ is either a public 
key for an implicit account (in case of a simple transfer)  or a
public hash for a smart contract (for an invocation), \PARAMETER\ is
the argument passed to the smart contract, 
which is empty for a simple transfer, and \MTEZ\ is
the amount of tokens for the transaction fee. In originate,  \CODE\ is
the script of a smart contract and \INIT\ is the initial value of the
contract's storage. Each operation returns an \emph{operation hash}
\OPH, on which we can query the status of the operation.

The local node offers several ways to query the current state of
the blockchain.
Some \emph{query operators} are defined by the following grammar:
\begin{align*}
  \QOP & ::=
  % \GETCOUNTER\PUK  
  %        \mid
         \GETBALANCE{}%\ADDR
  \mid \GETSTATUS{}%\OPH
  \mid \GETSTORAGE{}%\PUH
  % \mid \GETCODE\PUH 
  % \mid \GETTYPE\PUH
         \mid \GETCONTRACT{}%\OPH
         \mid \dots
\end{align*}
We obtain the balance associated with an implicit account or a
contract by its public key or public hash, respectively; the status of
a submitted operation by its operation hash; the stored value of a
contract by its public hash; and the public hash of a contract by the
operation hash of its originating transaction.

@@

\begin{figure}[tp]
\begin{align*}
  \CONSTANT & ::= \INT \mid \FIX
              \mid \OPH \mid \PUH \mid\PUK \mid \CODE \mid
              \NTEZ \mid \SUNIT \mid \FALSE \mid \TRUE \\
  \STATUS & ::= \STATUSPENDING \mid \STATUSINCLUDING (\INT) \mid
            \STATUSTIMEOUT \\
  \ERROR &::= \ERRPRG \mid \ERRBAL \mid \ERRCOUNT \mid \ERRFEE \mid
           \ERRPUK \mid \ERRPUH \mid \ERRARG \mid \ERRINIT\\
	\EXPR & :: =  \CONSTANT \mid \STATUS \mid \ERROR \mid \VARIABLE \mid \lambda \VARIABLE. \EXPR\ 
	\mid \EXPR\EXPR  %\mid \FST\ \EXPR \mid \SND\ \EXPR 	%\mid \NI \EI
  \mid \EXPR\ \PLUS\ \EXPR\ \mid
  % \EXPR\ \MINUS\ \EXPR\ \mid
  \EXPR\ \EQUAL\ \EXPR\
  % \mid  \EXPR\ \LESS\ \EXPR
  \mid \EXPR\ \AND\ \EXPR\ \mid \EXPR\ \OR\ \EXPR\ \mid \NOT\ \EXPR
  \\& \mid (\EXPR,\EXPR)\mid \NIL\ \mid \CONS\ \EXPR\ \EXPR \mid \LEFT\ \EXPR \mid\RIGHT\
  \EXPR \mid \SOME\ \EXPR \mid \NONE
  \mid \MATCH\ \EXPR\ \WITH\ \PATTERN\to\EXPR \dots 
  \\& \mid \RAISE\ \EXPR \mid \TRY\ \EXPR\ \EXCEPT\ \EXPR \mid \CAST\EXPR\TYPE\TYPEU
  \\& \mid \QOP\  \EXPR \mid \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR
  \\& \mid \ORIGINATE\EXPR\EXPR\EXPR\EXPR\EXPR
  % \\&\mid \CONTRACT\ \TYPE\ \EXPR
  % \\& \mid \TOSTRING\ \EXPR \mid \FROMSTRING\ \TYPE\ \EXPR
  \\
  \PATTERN &::= \VARIABLE \mid (\PATTERN, \PATTERN) \mid \NIL \mid \CONS\ \PATTERN\ \PATTERN \mid \LEFT\ \PATTERN \mid\RIGHT\
             \PATTERN \mid \SOME\ \PATTERN \mid \NONE \\
            &\mid \FALSE \mid \TRUE \mid \STATUS \mid \ERROR
  % \\
  % \QOP & ::= \GETBALANCE{} \mid \GETSTATUS{} \mid \GETSTORAGE{} \mid \GETCODE{} \mid \GETCOUNTER{}
  %        \mid \GETTYPE{}
  \\[2ex]
  \TYPE, \TYPEU & ::=
                  \TPUH \mid
                  \TPUK \mid
                  \TADDR \mid 
                  \TCONTRACT\ \TYPE\ \TYPEU \mid
                  \TCODE\ \TYPE\ \TYPEU \mid
                  \TOPH\ \TYPE\ \TYPEU \mid
                  \TSTATUS \mid \TEXCEPTION \mid \TTEZ \\
  & \mid \TNO \mid \TINT \mid \TUNIT \mid \TBOOL \mid \TSTRING \mid \TYPE\to\TYPEU \mid \TPAIR\ \TYPE\ \TYPEU \mid \TLIST\ \TYPE
    \mid \TSUM\ \TYPE\ \TYPEU \mid \TOPTION\ \TYPE 
\end{align*}
  \caption{Syntax of expressions, \EXPR, and types, \TYPE}
  \label{fig:syntax-expressions}
\end{figure}
All queries and operations are started by programs, i.e., closed
expressions of type unit that run on the
local node. Each program can send transactions on behalf of any
account on the local node.  Figure~\ref{fig:syntax-expressions} defines the syntax of
expressions, which describe the usual constructs of call-by-value PCF
with sum, product, list, and option types, exceptions and fixpoint.
Pattern matching is the only means to decompose values, cf.\ \PATTERN.

Moreover, there are
domain-specific primitive types and constants \CONSTANT\ to support blockchain
operations, as well as several exceptional values collected in \ERROR.
There is syntax to initiate transfers and to originate contracts as
well as for the queries. Finally, there is a type cast
$\CAST\EXPR\TYPE\TYPEU$, which we describe after discussing types. An
implementation provides all of these types and operations via a library API.

Types (also in Figure~\ref{fig:syntax-expressions}) comprise a
collection of standard base types as well as 
functions, pairs, lists, sums, and option types. These types are
chosen to match with built-in types of Michelson. 
In addition, there
are domain specific types of public hashes $\TPUH$ and public keys $\TPUK$ subsumed by
a type of addresses $\TADDR$. $\TCONTRACT\ \TYPE\ \TYPEU$ is the type of
a contract with parameter type $\TYPE$ and storage type $\TYPEU$. $\CODE\ \TYPE\ \TYPEU$
indicates a Michelson program with parameter type $\TYPE$ and storage
type $\TYPEU$. Tezos tokens have type $\TTEZ$. The type $\TOPH\ \TYPE\
\TYPEU$ signifies operation hashes
returned by blockchain operations. The parameters of the hash carry
the types when originating a contract. Otherwise, they are set to the
irrelevant type $\TNO$. We take the liberty of omitting irrelevant
type parameters, that is, we write $\TOPH$ for $\TOPH\ \TNO\ \TNO$.
Querying the status of an operation
returns a value of type $\TSTATUS$. Exceptions have type
$\TEXCEPTION$.

@@

Type casts are only applicable to certain pairs of
types governed by a relation $\SubType$, which could also serve as a
subtyping relation. It is given by the axioms
$\inferrule{}{\TPUH \SubType \TADDR}$, $\inferrule{}{\TPUK \SubType
  \TADDR}$, and $\inferrule{}{\TCONTRACT\ \TYPE\ \TYPEU \SubType \TPUH}$.
A cast from $\TYPE$ to $\TYPEU$ is only allowed
if $\TYPE \SubType \TYPEU$ (upcast) or $\TYPEU \SubType \TYPE$
(downcast). Upcasts always succeed, but downcasts may fail at run time.
In particular,  public hashes and public keys can both stand for addresses. Moreover, a smart
contract with parameter type $\TYPE$ is represented by its public hash
at run time. The corresponding downcast must check whether the public
hash is valid and has the expected parameter and storage type.

@@

    \inferrule{}{\JTypeExpr\TEnv\OPH{\TOPH\ \TYPE\ \TYPEU}}

@@

    \inferrule{
      \JTypeExpr\TEnv\EXPR\TYPE \\
      \TYPE \SubType \TYPEU \vee \TYPEU \SubType \TYPE
    }{
      \JTypeExpr\TEnv{\CAST\EXPR\TYPE\TYPEU}\TYPEU
    }
  \end{mathpar}
  \caption{Typing rules for expressions (excerpt)}
  \label{fig:typing-expressions}
\end{figure}
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{
      \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
      \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
      \JTypeExpr\TEnv{\EXPR_3}\TPUK \\
      % \JTypeExpr\TEnv{\EXPR_3}{\TCONTRACT\ \TUNIT} \\
      \JTypeExpr\TEnv{\EXPR_4}\TUNIT \\
      \JTypeExpr\TEnv{\EXPR_5}\TTEZ }{
      \JTypeExpr\TEnv{\TRANSFER[\EXPR_4]{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_5}}\TOPH\
      \TNO\ \TNO
    }

    \inferrule{
      \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
      \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
      \JTypeExpr\TEnv{\EXPR_3}\TCONTRACT\ \TYPE_p\ \TYPE_s \\
      \JTypeExpr\TEnv{\EXPR_4}{\TYPE_p} \\
      \JTypeExpr\TEnv{\EXPR_5}\TTEZ }{
      \JTypeExpr\TEnv{\TRANSFER[\EXPR_4]{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_5}}{\TOPH\
      \TNO\ \TNO}
    }

    \inferrule{
      \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
      \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
      \JTypeExpr\TEnv{\EXPR_3}{\TCODE\ \TYPE_p\ \TYPE_s} \\
      \JTypeExpr\TEnv{\EXPR_4}\TYPE_s \\ % really \TSTRING \\
      \JTypeExpr\TEnv{\EXPR_5}\TTEZ }{
      \JTypeExpr\TEnv{\ORIGINATE{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_4}{\EXPR_5}}{\TOPH\
      \TYPE_p\ \TYPE_s}
    }
  \end{mathpar}
\begin{mathpar}
  % \inferrule{
  %   \JTypeExpr\TEnv\EXPR\TPUK
  % }{
  %   \JTypeExpr\TEnv {\GETCOUNTER\EXPR}\TINT
  % }
  %   
  \inferrule{
    \JTypeExpr\TEnv\EXPR\TADDR
  }{
    \JTypeExpr\TEnv {\GETBALANCE\EXPR}\TTEZ
  }

  \inferrule{
    \JTypeExpr\TEnv\EXPR{\TOPH\ \TYPE\ \TYPEU}
  }{
    \JTypeExpr\TEnv {\GETSTATUS\EXPR}\TSTATUS
  }

  \inferrule{
    \JTypeExpr\TEnv\EXPR{\TCONTRACT\ \TYPE_p\ \TYPE_s}
  }{
    \JTypeExpr\TEnv {\GETSTORAGE\EXPR}\TYPE_s
  }

  % \inferrule{
  %   \JTypeExpr\TEnv\EXPR{\CONTRACT\ \TYPE}
  % }{
  %   \JTypeExpr\TEnv {\GETCODE\EXPR}?
  % }
  %   
  % \inferrule{
  %   \JTypeExpr\TEnv\EXPR{\CONTRACT\ \TYPE}
  % }{
  %   \JTypeExpr\TEnv {\GETTYPE\EXPR}?
  % }
  %   
  \inferrule{
    \JTypeExpr\TEnv\EXPR{\TOPH\ \TYPE\ \TYPEU}
  }{
    \JTypeExpr\TEnv {\GETCONTRACT\EXPR}{\TCONTRACT\ \TYPE\ \TYPEU}
  }
\end{mathpar}
  \caption{Typing rules for blockchain operations and queries}
  \label{fig:typing-blockchain-operations}
\end{figure}
% Generic rule for transfers
% \begin{mathpar}
%  \inferrule{
%    \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
%    \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
%    \JTypeExpr\TEnv{\EXPR_3}\TADDR \\
%    \JTypeExpr\TEnv{\EXPR_4}\TSTRING \\
%    \JTypeExpr\TEnv{\EXPR_5}\TTEZ
%  }{
%    \JTypeExpr\TEnv{\TRANSFER[\EXPR_4]{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_5}}\TOPH
%  }
% \end{mathpar}

% Conversion to and from serialized storage; deserialization may fail
% with an exception
% \begin{mathpar}
%   \inferrule{
%     \JTypeExpr\TEnv\EXPR\TYPE \\
%     \SERIALIZABLE (\TYPE)
%   }{
%     \JTypeExpr\TEnv{\TOSTRING\ \EXPR}\TSTRING
%   }

%   \inferrule{
%     \JTypeExpr\TEnv{\EXPR}\TSTRING
%   }{
%     \JTypeExpr\TEnv{\FROMSTRING\ \TYPE\ \EXPR}\TYPE
%   }
% \end{mathpar}
